Understanding the Lomb-Scargle Periodogram

### 2017-02-06 PracticalLombScargle.tex
Add summary of recommendations

Often this is exactly how the Lomb-Scargle periodogram is presented: as a
clean, well-defined procedure to generate a power spectrum and to
detect the periodic component in an unevenly-sampled dataset.
In practice, however, there are a number of subtle issues that must be
considered when applying a Lomb-Scargle analysis to real-world datasets.
Here are a few questions in particular that we might wish to ask about
the results in \fig{LINEAR-power}:
\begin{enumerate}
  \item What is the source of the pattern of multiple peaks revealed by the
    Lomb-Scargle Periodogram?
  \item What is the largest frequency (i.e. Nyquist-like limit) that such
    an analysis is sensitive to?
  \item How should we choose the spacing of the frequency grid for our
    periodogram?
  \item What assumptions is the Lomb-Scargle periodogram making about the
    shape of the unknown signal?
  \item How should we understand the uncertainty of the peak frequency?
\end{enumerate}
Quantitative answers to these sorts of questions are discussed in various
textbooks and review papers, but I have not come across any single concise
reference that gives a good intuition for how to think about such questions.
This paper seeks to fill that gap, and provide a practical,
just-technical-enough guide to the effective use of the Lomb-Scargle method
for detection and characterization of periodic signals.
This paper does not seek a complete or rigorous treatment of the mathematics
involved, but rather seeks to develop the intuition of {\it how} to think
about these questions.

@@

This paper has been a conceptual tour of the Lomb-Scargle periodogram, from
its roots in Fourier analysis, to its equivalence with special cases of
periodic analysis based on least squares model fitting and Bayesian
analysis. I will finish here with a brief summary of the practical
recommendations contained in these pages, along with a somewhat iconoclastic
post-script for further thought.

\subsection{Summary of Recommendations}

\todo{Flesh these out}

\begin{itemize}
\item choose your frequency range \& grid spacing carefully, so as to not
  miss relevant peaks.
\item always compute the window periodogram first, so that you know what alias
  structures to expect in the periodogram for your data.
\item always use the floating-mean model---it has many benefits with
  virtually zero down-sides.
\item the highest peak of the periodogram rarely tells the entire story.
  Always check for aliases that are expected from window structure
  ($f = f_{max} \pm n\delta f$) and from non-sinusoidal signals
  ($f = f_{max} / m$).
\item avoid Bayesian methods unless you are very comfortable with the
  assumptions they make about the form of your signal.
\item avoid peak widths as a proxy for uncertainty unless you are quite
  certain that your analysis has found the correct peak.
\item avoid false alarm probabilities as a proxy for uncertainty
  unless you make very clear {\it exactly} what kind of error they measure.
  Estimate the FAP via bootstrap if computationally feasible, and otherwise
  keep in mind the failures of available estimates.
\item in the context of a time-domain survey, always use simulated observations
  to evaluate your rates of false positives, false negatives, and failures to
  identify the correct frequency.
\item in cases of strong aliasing, consider using Lomb-Scargle as a first pass,
  and then select among candidate frequencies with a more detailed model of
  the signals you expect.
\end{itemize}

\subsection{Postscript: Why Lomb-Scargle?}
After considering all of these practical aspects of the periodogram,
I think it is worth stepping back to
revisit the question of {\it why} astronomers lean toward the
Lomb-Scargle approach rather than the (in many ways simpler) Schuster
periodogram.
In particular, three decades after the entry of ``Lomb-Scargle'' into the
astronomy community's vernacular, there seems to be a well-developed
mythology surrounding the efficacy and efficiency of the Lomb-Scargle approach.
It's common to see statements or implications like the following:
\begin{itemize}
  \item {\it Myth: The Lomb-Scargle periodogram can be computed more efficiently
    than the Schuster periodogram.} Reality: computationally, the two
    are quite similar, and in fact the fastest Lomb-Scargle algorithm currently
    available is based on the Schuster periodogram computed via the
    the NFFT algorithm (see \sect{algorithmic-considerations}).
  \item {\it Myth: The Lomb-Scargle periodogram is faster than a direct
    least squares periodogram because it avoids explicitly solving for
    model coefficients.}
    Reality: model coefficients can be determined with not much extra
    computation \citep[see discussion in][]{ICVG2014}.
  \item {\it Myth: The Lomb-Scargle periodogram allows analytic computation of
    statistics for periodogram peaks.} Reality: while this is true at
    individual frequencies, it is not true for the more relevant issue of
    peak-value statistics, which must be either approximated or computed by
    bootstrap resampling (see \sect{false-alarm-probability})
  \item {\it Myth: The Lomb-Scargle periodogram corrects for aliasing due to
    sampling and leads to independent noise at each frequency.}
    Reality: for structured window functions, the Lomb-Scargle
    periodogram has the same partial aliasing issues as the Schuster
    periodogram, including spurious peaks due to partial aliasing, and
    highly correlated periodogram errors (see \sect{failure-modes}).
  \item {\it Myth: Bayesian analysis shows that Lomb-Scargle is the optimal
    statistic for detecting periodic signals in data.} Reality: it
    shows that Lomb-Scargle is the optimal statistic for
    {\it fitting a sinusoid to data}, which is not the same as saying it
    is optimal for finding a generic, non-sinusoidal signal
    (see \sect{bayesian-periodograms}).
\end{itemize}
With these myths debunked, what advantage remains for Lomb-Scargle?
What would we lose if we instead used a Schuster periodogram,
estimating uncertainty, significance, and false alarm probability by
resampling and simulation, as we must for Lomb-Scargle itself?

@@

One clear advantage of both Lomb-Scargle and Schuster periodograms is that
they are quite fast to compute compared to more sophisticated approaches to
periodic analysis based on more constrained generative models or Bayesian
non-parametric models, and in the context of large time-domain surveys can
provide rough periodicity detections and catalog parameters in near real-time.

@@

{\it Acknowledgments:} I am grateful to Jeff Scargle, Max Mahlke,
Zeljko Ivezic, and Andy Connolly for helpful feedback on early drafts
of this paper.
This work was supported by the University of Washington eScience Institute,
with funding from the Gordon and Betty Moore Foundation, the Alfred P. Sloan
Foundation, and the Washington Research Foundation.

>>>>

Often this is exactly how the Lomb-Scargle periodogram is presented: as a
clean, well-defined procedure to generate a power spectrum and to
detect the periodic component in an unevenly-sampled dataset.
In practice, however, there are a number of subtle issues that must be
considered when applying a Lomb-Scargle analysis to real-world datasets.
Here are a few questions in particular that we might wish to ask about
the results in \fig{LINEAR-power}:
\begin{enumerate}
  \item What is the source of the pattern of multiple peaks revealed by the
    Lomb-Scargle Periodogram?
  \item What is the largest frequency (i.e. Nyquist-like limit) that such
    an analysis is sensitive to?
  \item How should we choose the spacing of the frequency grid for our
    periodogram?
  \item What assumptions does the Lomb-Scargle approach make regarding the
    shape of the unknown signal?
  \item How should we understand and report the uncertainty of the
    determined frequency?
\end{enumerate}
Quantitative treatments of these sorts of questions are presented in various
textbooks and review papers, but I have not come across any single concise
reference that gives a good intuition for how to think about such questions.
This paper seeks to fill that gap, and provide a practical,
just-technical-enough guide to the effective use of the Lomb-Scargle method
for detection and characterization of periodic signals.
This paper does not seek a complete or rigorous treatment of the mathematics
involved, but rather seeks to develop the intuition of {\it how} to think
about these questions.

@@

This paper has been a conceptual tour of the Lomb-Scargle periodogram, from
its roots in Fourier analysis, to its equivalence with special cases of
periodic analysis based on least squares model fitting and Bayesian
analysis. I will finish here with a brief summary of the practical
recommendations contained in these pages, along with a somewhat opinionated
and iconoclastic post-script for further thought.

\subsection{Summary of Recommendations}
\sectlabel{summary}
The previous pages contain a large amount of background and advice for working
with the Lomb-Scargle periodogram.
Here I will quickly summarize the steps you should take when applying this
algorithm to a dataset:

\begin{enumerate}
  \item Choose an appropriate frequency grid: the minimum can be set to zero,
    the maximum set based on the precision of the time measurements
    (\sect{}), and the grid spacing set based on the temporal baseline of
    the data (\sect{}) so as to not sample too coarsely around peaks.
    If this grid size is computationally intractable, reduce the maximum
    frequency based on what kinds of signals you are looking for (\sect{}).
  \item Compute the window transform using the Lomb-Scargle periodogram,
    by substituting $g_n=1$ for each $t_n$ and making sure not to pre-center
    the data or use a floating-mean model (\sect{}). Examine this window
    function for dominant features, such as daily or annual aliases (cf.
    \fig{}) or Nyquist-like limits (cf. \fig{}).
  \item Compute the periodogram for your data. You should always use the
    floating-mean model (\sect{}), as it produces more robust periodograms
    and has few if any disadvantages.
  \item Plot the periodogram, and identify any patterns that may be caused
    by features you observed in the window function power. Plot reference
    lines showing several False Alarm Probability levels to understand whether
    your periodogram peaks are significant enough to be labeled detections:
    use the Baluev method (\sect{})
    or the bootstrap method (\sect{}) 
    if it is computationally feasible.
    Keep in mind exactly what the False Alarm Probability measures, and
    avoid the temptation to misinterpret it (\sect{}).
  \item If the window function shows strong aliasing, locate the expected
    multiple maxima and plot the phased light curve at each
    (cf. \fig{}).If there is indication
    that the sinusoidal model under-fits the data (cf. \fig{}) then consider
    re-fitting with a multi-term Fourier model (\sect{}).
  \item If you have prior knowledge of the shape of light curves you are
    trying to detect, consider using more complex models to choose between
    multiple peaks in the periodogram (\sect{}). This type of refinement can
    be quite useful in building automated pipelines for period fitting,
    especially in cases where the window aliasing is strong.
  \item If you are building an automated pipeline based on Lomb-Scargle for
    use in a survey, consider injecting known signals into the pipeline to
    measure your detection efficiency as a function of object type, brightness,
    and other relevant characteristics (\sect{})
\end{enumerate}

This list is certainly not comprehensive for all uses of the periodogram, but
it should serve as a gentle reminder of the kinds of issues you should keep
in mind when using the method in practice.

\subsection{Other Measures of Periodicity}
\todo{fill this in}

\subsection{Postscript: Why Lomb-Scargle?}
\sectlabel{postscript}
After considering all of these practical aspects of the periodogram,
I think it is worth stepping back to
revisit the question of {\it why} astronomers lean toward the
Lomb-Scargle approach rather than the (in many ways simpler) Schuster
periodogram.

%One clear advantage retained by both the Lomb-Scargle and Schuster periodograms
%is that they are quite efficient compared with more sophisticated approaches
%to periodic analysis based on more constrained generative models or Bayesian
%non-parametric models, and in the context of large time-domain surveys can
%provide rough periodicity detections and catalog parameters in near real-time.

In particular, three decades after the entry of ``Lomb-Scargle'' into the
astronomy community's vernacular, there seems to be a well-developed
mythology surrounding the efficacy and efficiency of the Lomb-Scargle approach.
It's common to see statements or implications like the following:
\begin{itemize}
  \item {\it Myth: The Lomb-Scargle periodogram can be computed more efficiently
    than the Schuster periodogram.} Reality: computationally, the two
    are quite similar, and in fact the fastest Lomb-Scargle algorithm currently
    available is based on the Schuster periodogram computed via the
    the NFFT algorithm (see \sect{algorithmic-considerations}).
  \item {\it Myth: The Lomb-Scargle periodogram is faster than a direct
    least squares periodogram because it avoids explicitly solving for
    model coefficients.}
    Reality: model coefficients can be determined with little extra
    computation \citep[see the discussion in][]{ICVG2014}.
  \item {\it Myth: The Lomb-Scargle periodogram allows analytic computation of
    statistics for periodogram peaks.} Reality: while this is true at
    individual frequencies, it is not true for the more relevant issue of
    peak-value statistics, which must be either approximated or computed by
    bootstrap resampling (see \sect{false-alarm-probability})
  \item {\it Myth: The Lomb-Scargle periodogram corrects for aliasing due to
    sampling and leads to independent noise at each frequency.}
    Reality: for structured window functions, the Lomb-Scargle
    periodogram has the same window-driven issues as the Schuster
    periodogram, including spurious peaks due to partial aliasing, and
    highly correlated periodogram errors (see \sect{failure-modes}).
  \item {\it Myth: Bayesian analysis shows that Lomb-Scargle is the optimal
    statistic for detecting periodic signals in data.} Reality: Bayesian
    analysis shows that Lomb-Scargle is the optimal statistic for
    {\it fitting a sinusoid to data}, which is not the same as saying it
    is optimal for finding the frequency of a generic, potentially
    non-sinusoidal signal (see \sect{bayesian-periodograms}).
\end{itemize}
With these myths debunked, what advantage remains for Lomb-Scargle?
What would we lose if we instead used a Schuster periodogram,
estimating uncertainty, significance, and false alarm probability by
resampling and simulation, as we must for Lomb-Scargle itself?

@@

@@

\acknowledgments
{\it Acknowledgments:} I am grateful to Jeff Scargle, Max Mahlke,
Zeljko Ivezic, and Andy Connolly for helpful feedback on early drafts
of this paper.
This work was supported by the University of Washington eScience Institute,
with funding from the Gordon and Betty Moore Foundation, the Alfred P. Sloan
Foundation, and the Washington Research Foundation.
